---
title: Constructing a DAG for the impact of DOC on Anoxic Duration Under Ice 
author: "Katie Gannon"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Setting Up R Environment 

```{r cars, echo= FALSE, warning = FALSE}
library(ggdag)
require(knitr)
library(dagitty)
require(tidyr)
require(ddplyr)
library(tidyverse)
```

# Project Overview
<span style="color:purple;">
*Katie Comment:* I study lakes and for one of my projects I am interested in the processes happening under lake ice in winter. I will be using this exercise to explore how the amount of organic matter (measured as the concentration of dissolved organic carbon or DOC) in the water of a lake in the fall impacts the duration of anoxia (no oxygen) in bottom waters of mountain lakes. 
</span> 


# Create DAG

```{r}

# Create DAG  

    DAG_oxygen_under_ice <- dagify(
      
      #Identify Relatiobships 
      anox_duration ~ O2_at_Freeze + Ice_Duration + Rate_O2_Drawdown,
      Rate_O2_Drawdown ~ ER_under_ice + GPP_under_ice, 
      ER_under_ice ~ Freezing_Degree_Days + DOC_at_Freeze,
      DOC_at_Freeze ~ Terrestrial_DOC_Inputs + Summer_GPP, 
      GPP_under_ice ~ Summer_GPP + Light_Availability,
      Light_Availability ~ Snow_Accumulation + Ice_Thickness + Lake + Year,
      Ice_Thickness ~ Freezing_Degree_Days,
      Ice_Duration ~ Freezing_Degree_Days + Snow_Accumulation, 
      O2_at_Freeze ~ Summer_GPP + Lake + Fall_Mixing,
      Summer_GPP~ Terrestrial_DOC_Inputs + Lake + Year,
      Freezing_Degree_Days ~ Lake + Year, 
      Snow_Accumulation ~ Lake + Year,
      Terrestrial_DOC_Inputs ~ Lake + Year, 
      Fall_Mixing ~ Lake + Year, 
      
      # Specify exposure and outcome 
      exposure = "Snow_Accumulation", 
      outcome = "anox_duration",
      
      # Label 
      labels = c(outcome = "Annoxic Duration",
                 exposure = "DOC at Lake Freeze "))


# Visualize DAG
set.seed(124) #set the seed when plotting because otherwise the DAG shows up in different orientations each time!

ggdag_status(DAG_oxygen_under_ice,
             use_labels = "label",
             text = TRUE,
              text_col = "black",
             label_alpha = 0.5) + theme_dag()

```

<span style="color:purple;">
*Katie Comment:* The main updates that I have made are adding Lake and Year variables to my DAG (because I expect both of those variables to impact both my exposure and my impact variables) and thinking about the timescale of my variables. I realized that there was s disconnect because I was comparing some variables that were rates, some variables that had discrete measurements at individual timepoints, and some variables that were seasonal totals or averages. Those time steps made it tricky as I tried to think about how to account for more climate and seasonal variables. To deal with this disconnect, I shifted all of my variables to be either an descrete measure before the season of interest or an average/max across the season. 
</span>

#Identify Paths 
This function takes a DAG, with a given “exposure” or "treatment variable" and an “outcome” and identifies open backdoors: confounding variables (common causes) that need to be controlled for. 

```{r, echo = TRUE}
#identify the open paths that need to be adjusted for 
paths(DAG_oxygen_under_ice)

#plot the open  paths (aka a mess of bubbles)
ggdag_paths(DAG_oxygen_under_ice, shadow = TRUE) #Also, do not forget to set the argument shadow = TRUE, so that the arrows from the adjusted nodes are included

```
<span style="color:purple;">
[*Katie Comment:* With these added parameters and timesteps I have a BUNCH of open paths. ]
</span>

# Identify Decendent Nodes 
```{r, echo = TRUE}
#you can identify "descendant nodes" 
ggdag_descendants(DAG_oxygen_under_ice, "DOC_at_Freeze") 
# this gives you all of the variables taht are direct decendents of (impacted by) "DOC at Freeze" 
```

# Identify and Visualize Covariates that need to be adjusted for 
```{r, echo = TRUE}
#identify the covariaters need to be adjusted for to meet the back-door criterion
adjustmentSets(DAG_oxygen_under_ice)

# Finally, you can also visulaize the variables that need to be adjusted for (which is also told to you by the adjustmentSets() function 
ggdag_adjustment_set(DAG_oxygen_under_ice, shadow = TRUE) 
#Also, do not forget to set the argument shadow = TRUE, so that the arrows from the adjusted nodes are included.
```

<span style="color:purple;">
[*Katie Comment:* In terms of looking at adjustment sets though it looks like I am in pretty good shape! Based on the data that I have I can pretty easily account for Lake, Year, and average summer GPP.  ]
</span>

#Check Conditional independeces
We can also check to see which variables are conditionally independences in the DAG using impliedConditionalIndependencies().
```{r, echo = TRUE}
# Identify the implied conditional dependencies
impliedConditionalIndependencies(DAG_oxygen_under_ice)
```

<span style="color:purple;">
[*Katie Comment:* I still need to dig through all of these conditional dependencies with a finetooth comb but on a cersory look I think that we are in pretty good shape.]
</span>






